### 2023-03-12 chapter0/chapter0.tex
Tweaked wording in outline
Split large paragraph on recognition of disorder and added citation to early example of conformational heterogeneity

%I. Despite significant increases in the accuracy of computational methods which predict disorder from sequence alone, the relationship between sequence and function in IDRs remains poorly understood, and functions are not readily associated with arbitrary IDRs.

@@

%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequences and structure was an active area of research for decades
%            a. Recent developments in machine learning have largely resolved this question in most cases, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.

@@

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, indicating they lacked defined structures and were ``disordered.'' Because these segments were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of proteins. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions. In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat. Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.

Over the following years, many studies investigated the structural and functional properties of IDRs. They found that although IDRs still have sequence-structure-function relationships, they play by a very different set of rules. These differences manifest at all three levels but are at first most easily understood in terms of structure. Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms. Thus, ``structured'' regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein. Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically fold into a small number of related structures. This does not imply these folded domains are completely rigid, though. At the molecular level, everything is in constant motion. For example, at room temperature an average water molecule moves at over 500 meters per second!\footnote{This value was derived using the Maxwell–Boltzmann distribution, which is a physical model of the speeds of particles in an ideal gas. Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, liquid water is so dense that it will collide with something after moving only a fraction of its own length. Likewise, in the cellular environment folded domains are buffeted by collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain. Thus, while folded domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.

>>>>

%I. Despite significant advances in predicting disorder from sequence alone and relating those predictions to specific conformational profiles, the relationship between sequence and function in IDRs remains poorly understood. Thus, functions are not readily associated with arbitrary IDRs.

@@

%        5. In practice, however, this is extremely difficult. Fully mapping the relationship between sequence and structure of folded domains was an active area of research for decades
%            a. Recent developments in machine learning have in many senses solved this problem, but still required an advanced computational model trained on nearly two hundred thousand experimentally determined structures.

@@

The success of structural methods at elucidating the molecular details of protein function cemented the view that function largely depends on the presence of a fixed structure. While scientists understood proteins were not completely rigid and could adopt a variety of related structures, many believed that functional proteins largely had a single dominant structure~\cite{Karush1950}. Despite its strength, exceptions to the structure-function paradigm were known. For example, elastin is a protein secreted by cells which allows tissues like skin or blood vessels to repeatedly expand and contract. It imparts this elasticity by forming networks of disordered chains that act like molecular springs. When a tissue experiences a force, the chains stretch to accommodate it. When the force is removed, the chains return to their random orientations, which reduces their end-to-end length and forces the tissue to return to its original shape~\cite{Vrhovski1998, Alberts2014}. However, as a result of this unique role in providing tissue elasticity, elastin's disorder was viewed as a specific adaptation rather than a general mechanism of protein function. In other cases, proteins had regions which returned undefined or highly variable atomic coordinates when analysed with structural techniques, indicating they lacked defined structures and were ``disordered.'' Because these segments were often short loops between structured regions, they were seen as ``linkers'' which facilitated the structure of the functional portions of proteins. By the early 2000s, though, enough exceptions had accumulated that scientists began to recognize that fully and partially disordered proteins were involved in many biological processes~\cite{Plaxco1997, Wright1999, Dunker2001}. Many of these examples were proteins which folded on binding to their targets, commonly other proteins. This mechanism was a departure from the prevailing model of interactions between biological macromolecules which required highly stable and complementary interfaces, like two puzzle pieces fitting together. As a result, scientists speculated that disorder was an adaptation that allowed proteins to efficiently relay and regulate signals by enabling interactions with many possible targets. Furthermore, the flexibility of disordered protein would permit environmental conditions to easily modulate these interactions.

In the following years, as the complete genomes of several scientifically important model organisms such as \textit{S. cerevisiae} (baker's yeast), \textit{C. elegans} (roundworm), and \textit{D. melanogaster} (fruit fly) were sequenced for the first time, researchers applied computational methods for predicting disorder to the proteins inferred from their genomes. They discovered that disorder is ubiquitous in eukaryotic\footnote{All life belongs to one of three categories, or domains. Two, Archaea and Bacteria, are all single-celled organisms with simple cellular structures. In contrast, the cells of members of Eukarya, \textit{i.e.} eukaryotes, are complex and contain substructures called organelles, among other differences. All animals, plants, and fungi are eukaryotes, though the domain includes many microorganisms as well.} organisms, with estimates of the fraction of proteins containing disordered segments of greater than 30 residues\footnote{The amino acids that compose the links of a protein chain are conventionally called residues to distinguish them from their related, but chemically distinct, free forms.} ranging between 28 and 63\%~\cite{Dunker2000, Ward2004}. For reference, though the lengths of proteins can vary dramatically, a ``typical'' protein contains on the order of a few hundred residues, so these regions can compose a significant fraction of a protein's length. Because these segments were disordered in their native state, \textit{i.e.} under normal operating conditions, such segments were termed ``intrinsically disordered regions'' (IDRs) to emphasize the disorder was not induced by exposure to chemicals or heat. Furthermore, while most proteins were predicted to contain a mixture of structured domains and IDRs, some ``intrinsically disordered proteins'' (IDPs) were entirely or almost entirely disordered. Thus, disorder was recognized as a pervasive but poorly understood feature of proteins.

Many studies investigated the structural and functional properties of IDRs over the following years. They found that although IDRs still have sequence-structure-function relationships, they play by a very different set of rules. These differences manifest at all three levels but are at first most easily understood in terms of structure. Strictly speaking, a protein's structure refers its three-dimensional arrangement of atoms. Thus, ``structured'' regions in proteins, often called domains\footnote{Though there are various overlapping definitions, domains typically refer to independently folding regions of a protein. Domains are also described as discrete functional or evolutionary elements since proteins may contain several domains which are connected by unstructured ``linker'' sequences.}, typically fold into a small number of related structures. This does not imply these folded domains are completely rigid, though. At the molecular level, everything is in constant motion. For example, at room temperature an average water molecule moves at over 500 meters per second!\footnote{This value was derived using the Maxwell–Boltzmann distribution, which is a physical model of the speeds of particles in an ideal gas. Clearly, water is not a gas at room temperature, so it should be considered a rough approximation.} However, liquid water is so dense that it will collide with something after moving only a fraction of its own length. Likewise, in the cellular environment folded domains are buffeted by collisions with water and other molecules, but they are constrained by the rigid bonds and interactions between amino acid residues in the chain. Thus, while folded domains can ``flex'' and ``breathe,'' these motions are minor variations on their overall structure.
